A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-03 podc_paper/algorithm.tex
working on proof of propagation
working on proof; some edits to intro to save space

\subsection{Overview}
We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process. 
Each tree node  stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
See Figure \ref{tournament} for an example.
In this section, we assume for simplicity that each node has an infinite blocks array.
Section \ref{reducing} describes how to replace the infinite array by a representation that uses bounded space.

To perform an operation on the queue, a process $P$ creates a block containing that single 
operation and appends it to the blocks array in $P$'s leaf.
Then, $P$ attempts to propagate the operation to each node along the path from that leaf to the root of the tree.
We shall define a total order on all operations that have been propagated to the root, which 
will serve as the linearization ordering of the operations.

@@

\begin{figure}
\input{implicit.pdf_t}
\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{tournament}.
The leaf blocks simply show the \fld{element} field.
Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
and \eleft\ and \eright\ fields are shown using arrows as before.
Root blocks also contain the additional \fld{size} field.
The \fld{super} field is not shown.}
\end{figure}

@@

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State let \var{new} be a new \typ{Block} \label{initNewBlock}
    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$\label{createEndLeft}
    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$\label{createEndRight}
	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}\label{createSumEnq}
	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}\label{createSumDeq}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLength}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}

>>>>

\subsection{Overview}
We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process. 
Each tree node  stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
See Figure \ref{tournament} for an example.
In this section, we use an infinite array of blocks in each node.
Section \ref{reducing} describes how to replace the infinite array by a representation with bounded space.

To perform an operation on the queue, a process $P$ appends a new block containing that  
operation to the blocks array in $P$'s leaf.
Then, $P$ attempts to propagate the operation to each node along the path from that leaf to the root of the tree.
We shall define a total order on all operations that have been propagated to the root, which 
will serve as the linearization ordering of the operations.

@@

\begin{figure}
\input{implicit.pdf_t}
\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{tournament}.
The leaf blocks simply show the \fld{element} field.
Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
and \eleft\ and \eright\ fields are shown using arrows as before.
Root blocks also have the additional \fld{size} field.
The \fld{super} field is not shown.}
\end{figure}

@@

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State let \var{new} be a new \typ{Block} \label{initNewBlock}
    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$\label{createEndLeft}
    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$\label{createEndRight}
	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}\label{createSumEnq}
	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}\label{createSumDeq}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLength}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}\label{testEmpty}
        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}

### 2023-01-03 podc_paper/analysis.tex
working on proof; some edits to intro to save space

\section{Analysis}
In this section, we analyze the number of \nf{CAS} invocations and the time complexity of our algorithm.
\begin{proposition}
An \nf{Enqueue} or \nf{Dequeue} operation does at most $14\log p$ \nf{CAS} operations.
\end{proposition}
\begin{proof}
  In each level of the tree  \nf{Refresh} is invoked at most two times, and every \nf{Refresh} invokes at most seven \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
\end{proof}

>>>>

\section{Analysis}
In this section, we analyze the number of \nf{CAS} invocations and the step complexity of our algorithm.

First, we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.

\begin{lemma}\label{blockSize}
Each block $B$ in each node contains at most one operation of each process.
If $c$ is the execution's maximum point contention, $B$ has at most $c$ direct~subblocks.
\end{lemma}
\begin{proof}
Suppose $B$ contains an operation of process $p$.
Let $op$ be the earliest operation by $p$ contained in $B$.
When $op$ terminates, $op$ is contained in $B$ by \Cref{lem::appendExactlyOnce}.
Thus, $B$ cannot contain any later operations by $p$, since $B$ is created before
those operations are invoked.

Let $t$ be the earliest termination of any operation contained in $B$.
By \Cref{lem::appendExactlyOnce}, $B$ is created before $t$, so all operations contained in $B$
are invoked before $t$.  Thus, they are all concurrently running at $t$, so there are at most
$c$ operations contained in $B$.
By definition, the direct subblocks of $B$ contain these $c$ operations, and each operation is contained
in exactly one of these subblocks, by Lemma \ref{lem::subblocksDistinct}.
By \Cref{blockNotEmpty}, each direct subblock of $B$ contains at least one operation,
so there are at most $c$ direct subblocks.
\end{proof}

\begin{proposition}
An \nf{Enqueue} or \nf{Dequeue} operation does at most $14\log p$ \nf{CAS} operations.
\end{proposition}
\begin{proof}
  In each level of the tree  \nf{Refresh} is invoked at most two times, and every \nf{Refresh} invokes at most seven \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
\end{proof}

### 2023-01-03 podc_paper/correctness.tex
working on proof; some edits to intro to save space

\begin{lemma}
Consider two consecutive terminating calls $R_1$, $R_2$ to \nf{Refresh}($v$) by the same process.
When $R_2$ terminates, $v$ contains all operations that $v$'s children contained when $R_1$ was invoked.
\end{lemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{cas} or terminates at line \ref{addOP}, the claim follows
from Lemma \ref{successfulRefresh}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{cas}.
Let $h_1$ and $h_2$ be the values $R_1$ and $R_2$ read from $v.\head$ at line \ref{readHead}.
By \Cref{lem::headInc}, $h_2>h_1$.
By \Cref{lem::headProgress}, $v.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
Since $R_2$ fails its \op{CAS} on $v.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
a successful \op{CAS} on $v.blocks[h_2]$ before $R_2$'s \op{CAS}.
$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $v.\head$ and the value of $v.\head$ is non-decreasing, by \Cref{nonDecreasingHead}.
Thus, all operations contained in $v$'s children when $R_1$ begins
are also contained in $v$'s children when $R_3$ later executes line \ref{readHead}.
By \Cref{successfulRefresh}, these operations are contained in $v$ when $R_3$ performs its successful \op{CAS},
which is before $R_2$ terminates.
\end{proof}

-------

\begin{corollary} \label{lem::propagateStep}
  $EST_{n\nf{.left}}^{_{\ref{firstRefresh}}t} \cup EST_{n\nf{.right}}^ {_{\ref{firstRefresh}}t} \subseteq EST_{n}^{t_{\ref{secondRefresh}}}$
\end{corollary}
\begin{proof}
  If the first \nf{Refresh} in line \ref{firstRefresh} returns \nf{true}, then by Corollary \ref{lem::prectrueRefresh} the claim holds. If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded, the claim still holds by Corollary \ref{lem::prectrueRefresh}. Otherwise, both failed, and the claim is implied by Lemma \ref{doubleRefresh}.
\end{proof}

Now we show that after \nf{Append($B$)} on a leaf finishes, the operation contained in $B$ will be established in \nf{root}.

\begin{corollary}\label{lem::append}
For $A=l\nf{.Append(}B\nf{)}$ we have ${ops(b)} \subseteq EST_{n}^{t^A}$ for each node $n$ in the path from $l$ to \nf{root}.
\end{corollary}
\begin{proof}
$A$ adds $B$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the parent of the leaf where it appended $B$. For every node $n$, $n.$\nf{Propagate} appends $B$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} until $n$ is \nf{root}.
\end{proof}

\begin{corollary} \label{lem::appendExactlyOnce}
 After $l$\nf{.Append($B$)} finishes, $B$ is a subblock of exactly one block in each node along the path from $l$ to the \nf{root}.
\end{corollary}
\begin{proof}
By the previous corollary and Lemma \ref{lem::noDuplicates} there is exactly one block in each node containing~$B$.
\end{proof}

\subsection{Correctness of GetEnqueue}

First, we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.

\begin{lemma}\label{oneOpinBlock}
Each block contains at most one operation of each process.
\end{lemma}
\begin{proof}
To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $P$ in block $B$ in node $n$. Without loss of generality $op_1$ is invoked earlier than $op_2$. Process $P$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins. By Corollary \ref{lem::appendExactlyOnce}, before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $P$'s leaf to the root. Since $b$ contains $op_2$, it must be created after $op_2$ is invoked. The fact that $op_2$\nf{.Append} is invoked after $op_1$\nf{.Append} terminated means that there is some block $B^\prime$ in $n$ before $B$ that contains $op_1$. The existence of $op_1$ in $B$ and $B^\prime$ contradicts Lemma~\ref{lem::noDuplicates}.
\end{proof}

\begin{lemma}\label{blockSize}
    Each block contains at most $c$ operations, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
\end{lemma}
\begin{proof}
There is a time that all the operations in a block are concurrent, because otherwise if there is an operation in a block that has ended before another operation in that block starts, then by Corollary \ref{lem::appendExactlyOnce} these two operations couldn't be in the same block. From the definition of $c$ we know at any time in the execution there cannot be more than $c$ concurrent operations, and from the previous lemma we know a process has at most one operation in a block, so there cannot be a block with more than $c$ operations.
\end{proof}

\begin{lemma}\label{subBlocksBound}
Each block has at most $c$ direct subblocks, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
\end{lemma}
\begin{proof}
From Definition \ref{def::ops} we know the operations in a block are the union of the operations in the direct subblocks of the block. We can see that each block appended to an internal node contains at least one operation due to the test on Line~\ref{addOP}. Also, blocks in the leaves contain only one \nf{Enqueue} or \nf{Dequeue} operation. By Lemma \ref{blockSize} each block in an internal node contains at most $c$ operations and each one of its direct subblocks has at least one operation, so by pigeonhole principle the number of direct subblocks in a block is at most $c$.
\end{proof}

>>>>

\begin{lemma}\label{lem::doubleRefresh}
Consider two consecutive terminating calls $R_1$, $R_2$ to \nf{Refresh}($v$) by the same process.
When $R_2$ terminates, $v$ contains all operations that $v$'s children contained when $R_1$ was invoked.
\end{lemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{cas} or terminates at line \ref{addOP}, the claim follows
from Lemma \ref{successfulRefresh}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{cas}.
Let $h_1$ and $h_2$ be the values $R_1$ and $R_2$ read from $v.\head$ at line \ref{readHead}.
By \Cref{lem::headInc}, $h_2>h_1$.
By \Cref{lem::headProgress}, $v.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
Since $R_2$ fails its \op{CAS} on $v.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
a successful \op{CAS} on $v.blocks[h_2]$ before $R_2$'s \op{CAS}.
$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $v.\head$ and the value of $v.\head$ is non-decreasing, by \Cref{nonDecreasingHead}.
Thus, all operations contained in $v$'s children when $R_1$ begins
are also contained in $v$'s children when $R_3$ later executes line \ref{readHead}.
By \Cref{successfulRefresh}, these operations are contained in $v$ when $R_3$ performs its successful \op{CAS},
which is before $R_2$ terminates.
\end{proof}

\begin{lemma} \label{lem::appendExactlyOnce}
When an \op{Append}($B$) terminates, $B$'s operation is contained in exactly one block in each node along the path from the process's leaf to the root.
\end{lemma}
\begin{proof}
The \op{Append} adds $B$ to the process's leaf and calls \op{Propagate}, which
performs a double \op{Refresh} on each node along the path from the leaf's parent to the root.
By \Cref{lem::doubleRefresh}, this ensures $B$'s operation is contained in each node along this path.
By \Cref{lem::noDuplicates}, it is contained in exactly one block in each node.
\end{proof}

\subsection{Correctness of GetEnqueue}
-------

### 2023-01-03 podc_paper/introduction.tex
working on proof; some edits to intro to save space

Many concurrent data structures that keep track of a set of elements also have an $\Omega(p)$ term in their step complexity, as observed by Ruppert \cite{Rup16}.
For example, lock-free lists \cite{FR04,Sha15}, stacks \cite{Tre86} and search trees \cite{EFHR14} 
have an $\Omega(c)$ term in their step complexity, where $c$ represents contention,
the number of processes that access the data structure concurrently (which can be $p$ in the worst case).
Attiya and Fouren \cite{DBLP:conf/opodis/AttiyaF17} proved 
that amortized $\Omega(c)$ steps per operation are indeed necessary
for CAS-based implementations of lock-free bag data structures, which provide operations
to insert an element or remove an arbitrary element (chosen non-deterministically).
Since a queue trivially implements a bag, this lower bound also holds for queues.
At first glance, this would seem to settle the question of the step complexity of lock-free queues.
However, the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
Thus, the lower bound could be stated more precisely as an amortized bound of $\Omega(\min(c,\log\log p))$ steps per operation.

In this paper, we exploit this loophole.  We show that it is in fact possible to implement a linearizable, lock-free queue
whose step complexity does not depend linearly on $p$.
Our implementation is the first to have step complexity that is polylogarithmic in $p$ and in $q$, the number of elements
stored in the queue.
For simplicity of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation.
Then, we show how to modify the implementation to use 
\here{$O(?)$ space} and \here{$O(?)$ steps per operation}.
Both implementations use single-word CAS instructions and reasonably-sized words.
Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions in an amortized sense.
For the space-bounded implementation, we unlink unneeded memory from our data structure.
We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

Our queue uses a tournament tree where each process has its own leaf.
Each process propagates its operations along the path from its leaf up to the root.
The root of the tree provides an ordering for all the operations,
and this ordering is used to linearize the operations and compute their responses.
\here{Either here or in related work section, talk about previous usage of tournament tree and how ours differs from it}
To ensure wait-freedom, we use a helping mechanism:  a process propagating its operation 
to a node $n$ helps propagate all operations from both of $n$'s children.
If we explicitly stored all operations in the nodes, this helping would become very costly.
Instead, we use an implicit representation of collections
of operations that allows for quickly merging two collections from the two children of a node,
and for quickly accessing any particular operation in the collection.
\here{Maybe say a little more about techniques used in the implementation, if space permits}

>>>>

Many concurrent data structures that keep track of a set of elements also have an $\Omega(p)$ term in their step complexity, as observed by Ruppert \cite{Rup16}.
For example, lock-free lists \cite{FR04,Sha15}, stacks \cite{Tre86} and search trees \cite{EFHR14} 
have an $\Omega(c)$ term in their step complexity, where $c$ represents contention,
the number of processes that access the data structure concurrently, which can be $p$ in the worst case.
Attiya and Fouren \cite{DBLP:conf/opodis/AttiyaF17} proved 
that amortized $\Omega(c)$ steps per operation are indeed necessary
for any CAS-based implementation of a lock-free bag data structure, which provide operations
to insert an element or remove an arbitrary element (chosen non-deterministically).
Since a queue trivially implements a bag, this lower bound also holds for queues.
At first glance, this would seem to settle the question of the step complexity of lock-free queues.
However, the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
Thus, the lower bound could be stated more precisely as an amortized bound of $\Omega(\min(c,\log\log p))$ steps per operation.

We exploit this loophole.  We show that it is, in fact, possible to implement a linearizable, wait-free queue
whose step complexity does not depend linearly on $p$.
Our implementation is the first to have step complexity that is polylogarithmic in $p$ and in $q$, the number of elements
stored in the queue.
For ease of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation.
Then, we show how to modify the implementation to use 
\here{$O(?)$ space} and \here{$O(?)$ steps per operation}.
Both implementations use single-word CAS and reasonably-sized words.
Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions, even in an amortized sense.
For the space-bounded implementation, we unlink unneeded memory from our data structure.
We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

Our queue uses a tournament tree where each process has its own leaf.
Each process propagates its operations along the path from its leaf up to the root.
The root of the tree provides an ordering for all the operations,
and this ordering is used to linearize the operations and compute their responses.
\here{Either here or in related work section, talk about previous usage of tournament tree and how ours differs from it}
To ensure wait-freedom, we use a helping mechanism:  a process propagating its operation 
to a node $n$ helps propagate all operations from both of $n$'s children.
If we explicitly stored all operations in the nodes, this helping would become very costly.
Instead, we use a novel implicit representation of collections
of operations that allows for quickly merging two collections from the two children of a node,
and for quickly accessing any particular operation in the collection.
\here{Maybe say a little more about techniques used in the implementation, if space permits}

### 2023-01-03 podc_paper/related.tex
working on proof; some edits to intro to save space

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} 
added Herlihy's helping
technique~\cite{10.1145/114005.102808} to the MS-queue to obtain
a wait-free queue.
Then, Kogan and Petrank \cite{10.1145/2370036.2145835} introduced the 
fast-path slow-path methodology for making data structures wait-free:
The fast path has good performance and the slow path guarantees termination.
They applied their methodology to combine the MS-queue (as the fast path)
with their previous wait-free queue (as the slow path).
Ramalhete and Correia \cite{RC17} added a different helping mechanism to the MS-queue.
Although these approaches can improve practical performance,
their amortized step complexity remains $\Omega(p)$. 
\Hossein{I get what this sentence means, but from a high overview i think it is inaccurate. Since the step complexity of MS-queue is infinite, so nothing can be added to it and a wait-free queue is always better than a lock-free queue in the worst-case.}
\Eric{I modified the wording.  I also added the material about fast-path slow-path here (since I had forgotten that paper also applies it to a queue), and fixed the paragraph about FPSP method below when discussin array-based queues.}

@@

Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} also used a linked list of (circular) arrays.
To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch\&increment instructions.
If livelock between enqueues and a racing dequeue prevent enqueues from claiming a spot,
the enqueues fall back on using a CAS to add a new array to the linked list, 
and the CAS retry problem reappears.
% I looked at the CRQ algorithm to implement each of the circular arrays.
% Here is an execution that causes livelock for 3 processes
% enqueue(A) reads t=0 and increments tail to 1
% enqueue(B) reads t=1 and increments tail to 2
% (*) dequeue reads h=0 and increments head to 1
% dequeue's CAS succeds changing array[0] to (1,R,bottom)
% dequeue reads t=2 at line 53
% enqueue(A) fails CAS on array[0]
% dequeue reads h=1 and increments head to 2
% dequeue's CAS succeeds changing array[1] to (1,R+1,bottom)
% enqueue(B) fails CAS on array[1]
% enqueue(A) reads t=2 and increments tail to 3
% enqueue(B) reads t=3 and increments tail to 4
% dequeue reads t=4 at line 53
% go back to (*) and repeat this sequence
This approach is similar to the fast-path slow-path methodology \cite{10.1145/2370036.2145835}.
Other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} have also used this methodology
on array-based queues.
In worst-case executions, where processes must execute the slow path,
these implementations also take $\Omega(p)$ steps per operation, 
due either to the CAS retry problem or helping mechanisms.

@@

\emph{Fetch\&Increment Objects.}
There are sublinear-time implementations of another object
with consensus number 2.
Two papers \cite{ERW12,10.1007/978-3-642-41527-2_20}
implemented fetch\&increment objects from LL/SC objects
using a polylogarithmic number of steps per operation.
Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
which uses a tournament tree to keep track of the fetch\&increments
that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.
However, our construction requires more intricate
data structures to represent sets of operations, since the state of a queue cannot easily be represented as succinctly
as the single-word state of a fetch\&increment object.

>>>>

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} 
used Herlihy's helping
technique~\cite{10.1145/114005.102808} to make the MS-queue
wait-free.
Then, they introduced the 
fast-path slow-path methodology \cite{10.1145/2370036.2145835} for making data structures wait-free:
the fast path has good performance and the slow path guarantees termination.
They applied their methodology to combine the MS-queue (as the fast path)
with their wait-free queue (as the slow path).
Ramalhete and Correia \cite{RC17} added a different helping mechanism to the MS-queue.
Although these approaches can perform well in practice,
the amortized step complexity remains~$\Omega(p)$. 
\Hossein{I get what this sentence means, but from a high overview i think it is inaccurate. Since the step complexity of MS-queue is infinite, so nothing can be added to it and a wait-free queue is always better than a lock-free queue in the worst-case.}
\Eric{I modified the wording.  I also added the material about fast-path slow-path here (since I had forgotten that paper also applies it to a queue), and fixed the paragraph about FPSP method below when discussing array-based queues.}

@@

Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} also used a linked list of (circular) arrays.
To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch\&increment instructions.
If livelock between enqueues and a racing dequeue prevent enqueues from claiming a spot,
the enqueues fall back on using a CAS to add a new array to the linked list, 
and the CAS retry problem reappears.
% I looked at the CRQ algorithm to implement each of the circular arrays.
% Here is an execution that causes livelock for 3 processes
% enqueue(A) reads t=0 and increments tail to 1
% enqueue(B) reads t=1 and increments tail to 2
% (*) dequeue reads h=0 and increments head to 1
% dequeue's CAS succeds changing array[0] to (1,R,bottom)
% dequeue reads t=2 at line 53
% enqueue(A) fails CAS on array[0]
% dequeue reads h=1 and increments head to 2
% dequeue's CAS succeeds changing array[1] to (1,R+1,bottom)
% enqueue(B) fails CAS on array[1]
% enqueue(A) reads t=2 and increments tail to 3
% enqueue(B) reads t=3 and increments tail to 4
% dequeue reads t=4 at line 53
% go back to (*) and repeat this sequence
This approach is similar to the fast-path slow-path methodology \cite{10.1145/2370036.2145835}.
Other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} also used this methodology
on array-based queues.
In worst-case executions that use the slow path,
they also take $\Omega(p)$ steps per operation, 
due either to the CAS retry problem or helping mechanisms.

@@

\emph{Fetch\&Increment Objects.}
There are sublinear-time implementations of another object
with consensus number 2.
Two papers \cite{ERW12,10.1007/978-3-642-41527-2_20}
implemented fetch\&increment objects from LL/SC objects
using a polylogarithmic number of steps per operation.
Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
which uses a tournament tree to keep track of the fetch\&increments
that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.
However, our construction requires more intricate
data structures to represent sets of operations, since a queue's state cannot be represented as succinctly
as the single-word state of a fetch\&increment object.
